Mental Rotation as Bayesian Quadrature

### 2013-05-31 man/nips-2013.tex
Abstract and referrences

\title{Mental Simulation as Bayesian Quadrature}
\author{Jessica B.~Hamrick and Thomas L.~Griffiths\\
  Department of Psychology\\
  University of California, Berkeley\\
  Berkeley, CA 94720\\
  \texttt{\{jhamrick,tom\_griffiths\}@berkeley.edu}}

@@

\begin{abstract}
\TODO{}
\end{abstract}

@@

One of the challenges of solving any computational problem is
determining how best to use the available computing resources. For
example, a computer can render complex graphics faster by recognizing
that this kind of computation should be carried out by a specialized
graphics processor. The same challenge arises in designing an
intelligent agent: how should the agent make best use of its computing
resources? Recent research on rational models of human cognition has
provided insight into the nature of the computational problems that
human beings need to solve (e.g. \cite{Chater:1999wp,Griffiths2012a}),
but leaves open the question of how people allocate their resources in
solving those problems. In this paper, we take a step towards
addressing this question, applying rational analysis (in the spirit of
\cite{Marr:1983to,anderson90} \TODO{shepard?}) to one aspect of human
metacognition: the use of mental simulation.

@@

Recent work in cognitive science has shown how problems of allocating
cognitive resources to solving computational problems can be analyzed
using the methods of statistical decision theory
\cite{Lieder:2012wg,Vul:2009wy}. We apply this ``rational
metacognition'' approach to the problem of mental rotation. We
investigate several computational solutions to this task and
qualitatively compare them to human mental rotation performance. In
particular, we argue that performing mental rotation can be framed as
an integration problem, with the direction of rotation becoming a
statistical design problem (or in machine learning parlance, a problem
of active learning). We show that recent work on methods for Bayesian
quadrature \cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}
provides a way to solve this problem, outperforming several simpler
heuristics for determining both the direction and extent of rotation.

@@

For many years, this idea was contested, with some researchers arguing
the underlying cognitive processes were not visual in nature
(e.g. \cite{Pylyshyn1981}). In particular, \cite{Anderson1978} proved
that the mental imagery debate could not be resolved on the basis of
response time data alone. In recent years, however, a significant
amount of work has investigated the neural underpinnings of the mental
rotation phenomena, arguing that the results of brain imaging studies
support the existence of visual mental rotation \cite{Kosslyn1988,
  Kosslyn:2009tj}.

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/stimuli_shapes.pdf}
    \vspace{0pt}
    \caption{Example stimuli}
    \label{fig:stimuli}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/likelihood_function.pdf}
    \caption{Likelihood function}
    \label{fig:likelihood}
  \end{subfigure}
  \caption{\textbf{Rotated shapes and their similarity.}  \textbf{(a)}
    An example stimulus in which the shapes differ only by a
    rotation. All stimuli consist of three to six vertices centered
    around the origin, and edges which create a closed loop from the
    vertices. \textbf{(b)} The approximate likelihood (similarity) of
    $X_b$ given $X_R$, where $X_R$ is a rotation of $X_a$ by the angle
    $R$. The true angle of rotation between $X_a$ and $X_b$ is at
    $\frac{2\pi}{3}$, corresponding to the global maximum in the
    likelihood function.}
  \label{fig:shapes}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/li_regression.pdf}
    \caption{\Naive{} model}
    \label{fig:li}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/vm_regression.pdf}
    \caption{Parametric model}
    \label{fig:vm}
  \end{subfigure}
  \caption{\textbf{Simple methods of estimating $S$.} In both cases,
    hill-climbing search is used until a maxima is found (in this
    case, at approximately $\frac{\pi}{6}$). The sampled points
    $\mathbf{R}$ (red circles) are then used to estimate $S$ (black
    lines are the true $S$, red lines are the estimate). \textbf{(a)}
    Linear interpolation. The overall estimate of $Z$ here will be too
    low, but certain angles will have a disproportionate contribution
    (e.g. between $\frac{\pi}{4}$ and $\frac{3\pi}{4}$. \textbf{(b)}
    Best fit of scaled Von Mises PDF parameters. As in (a), $Z$ will
    be underestimated. However, the fit around the maximum near
    $\frac{\pi}{6}$ is much more accurate.}
  \label{fig:simple-models}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{../figures/bq_regression.pdf}
  \caption{\textbf{Nonparametric model.} Each panel shows one step of
    the Bayesian Quadrature regression. Upper left: the original
    Gaussian Process (GP) regression for $S$. Lower left: the GP
    regression for $\log(S+1)$. Lower right: the GP regression for
    $\Delta_c=\mu_{\log S} - \log \mu_S$. Upper right: the adjusted
    regression for $S$, where the mean is equal to
    $\mu_S(1+\mu_{\Delta_c})$. The model uses this final estimate to
    compute $Z$ and will continue rotating until the variance of $Z$
    is low enough that a hypothesis may be accepted. This method
    allows the model to avoid local maxima such as the one near
    $\frac{\pi}{6}$, which causes trouble for the \naive{} and
    parametric models in Figure \ref{fig:simple-models}.}
  \label{fig:bq}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{../figures/model_rotations.pdf}
  \caption{\textbf{Model rotations.} Top: each subplot shows the
    correspondence between the true angle of rotation ($R$) for
    ``same'' pairs and the amount of rotation performed by the
    model. Bottom: each subplot shows the models' mean rotations over
    stimuli pairs that were judged correctly. Black dots correspond to
    ``same'' pairs, and blue lines correspond to ``different'' pairs
    (for which the mean rotation is constant across true rotations, as
    the true rotation is undefined). Error bars/shaded regions
    indicate one standard deviation, and the dotted lines indicate the
    least-squares fit to the ``same'' pairs.}
  \label{fig:rotations}
\end{figure}

@@

The \naive{} model's response error rate was $\NaiveME{}$, which is
better than chance (equivalent to guessing randomly,
i.e. $\ME{}=0.5$). \TODO{binomial test?} Closer inspection reveals
that much of this comes from ``different'' pairs ($\NaiveMEdiff{}$),
where the asymmetric linear interpolation may give an overestimate of
$Z$ (e.g. Figure \ref{fig:li}).  This intuition is confirmed by the
model's accuracy in estimating $Z$, which was $\NaiveMSE{}$. As shown
in the left panel of Figure \ref{fig:accuracy}, it overestimated $Z$
for nearly all stimuli pairs. \TODO{how many?}

@@

The parametric model's error in determining whether the objects were
identical was $\VMME{}$, making more errors on ``same'' pairs
($\VMMEsame{}$) than ``different'' pairs ($\VMMEdiff{}$). \TODO{is
  this significant?} Rather than overestimating as with the \naive{}
model, however, the parametric model largely underestimated $Z$. This
is unsurprising: because the Von Mises distribution only has a single
peak, it necessarily underestimates $Z$ for multimodal similarity
functions. Overall, the parametric model had an error rate of
$\VMMSE{}$ in estimating $Z$, which was significantly more accurate
than that of the \naive{} model \TODO{test}.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{../figures/Z_accuracy.pdf}
  \caption{\textbf{Accuracy in estimating $Z$.} Each subplot shows the
    true (``gold standard'') value of $Z$ vs. the value estimated by
    the model. Black dotted lines indicate a perfect 1:1
    correspondence between the true and estimated values. The \naive{}
    model (left panel) tends to overestimate. The parametric model
    (center panel) is very accurate for about half the ``same'' pairs,
    and severely underestimates the rest. The nonparametric model
    (right panel) maintains a decent correspondence with the true
    $Z$.}
  \label{fig:accuracy}
\end{figure}

@@

The nonparametric model was much more accurate in choosing the correct
hypothesis than the other two models ($\BQME{}$) \TODO{test}. We note
that this is very close to the 3.2\% error rate reported by
\cite{Shepard1971}, though further experimentation is necessary to
determine whether the types of errors people make align with those of
the model.

@@

% One option for improving the performance of the simple models would be
% to fit a threshold value, below which local maxima would be
% ignored. However, this strategy would be rather brittle, for if the
% distribution of shapes changed, the threshold would have to be
% re-learned. It is possible that people exhibit this behavior, but we
% cannot make any assumptions one way or the other without empirical
% data.

% A further strength of the nonparametric model over the other models,
% however, is that it is likely to generalize well to three
% dimensions. One aspect of \cite{Osborne:2012tm} which we have not yet
% explored is their main contribution of \textit{active sampling}, in
% which samples are iteratively selected to maximally decrease the
% expected variance of $Z$. Due to the sequential constraint of mental
% rotation, this strategy is not particularly useful in two dimensions,
% as there are only every two directions in which to rotate. In three
% dimensions, however, there are an infinite number of directions that
% could be chosen after every step.

>>>>

\title{Mental Rotation as Bayesian Quadrature}
\author{Jessica B.~Hamrick and Thomas L.~Griffiths\\
  Department of Psychology\\
  University of California, Berkeley\\
  Berkeley, CA 94720\\
  \texttt{\{jhamrick,tom\_griffiths\}@berkeley.edu}}

@@

\begin{abstract}
  Given a computationl resource--for example, the ability to visualize
  an object rotating--how do you best make use of it? We explored how
  mental simulation should used in the classic psychological task of
  determining if two images depict the same object in different
  orientations. We compared three models on this mental rotation task,
  and found that a model based on an optimal experiment design for
  Bayesian quadrature is objectively more accurate and qualitatively
  more consistent with classic behavioral data than the other models
  we evaluated. We suggest that rational models which adaptively
  exploit available resources are promising in their ability to
  characterize metacognitive processes like mental simulation.
\end{abstract}

@@

One of the challenges of solving any computational problem is
determining how best to use the available computing resources. For
example, a computer can render complex graphics faster by recognizing
that this kind of computation should be carried out by a specialized
graphics processor. The same challenge arises in designing an
intelligent agent: how should the agent make best use of its computing
resources? Recent research on rational models of human cognition has
provided insight into the nature of the computational problems that
human beings need to solve (e.g. \cite{Chater:1999wp,Griffiths2012a}),
but leaves open the question of how people allocate their resources in
solving those problems. In this paper, we take a step towards
addressing this question, applying rational analysis (in the spirit of
\cite{Marr:1983to,anderson90,Shepard:1987tt}) to one aspect of human
metacognition: the use of mental simulation.

@@

Recent work in cognitive science has shown how problems of allocating
cognitive resources to solving computational problems can be analyzed
using the methods of statistical decision theory
\cite{Lieder:2012wg,Vul:2009wy}. We apply this ``rational
metacognition'' approach to the problem of mental rotation. We
investigate several computational solutions to this task and
qualitatively compare them to human mental rotation performance. In
particular, we argue that performing mental rotation can be framed as
an integration problem, with the direction of rotation becoming a
problem of optimal experiment design (or in machine learning parlance,
of active learning). We show that recent work on methods for Bayesian
quadrature \cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}
provides a way to solve this problem, outperforming several simpler
heuristics for determining both the direction and extent of rotation.

@@

For many years, this idea was contested, with some researchers arguing
the underlying cognitive processes were not visual in nature
(e.g. \cite{Pylyshyn1981}). In particular, \cite{Anderson1978} proved
that the mental imagery debate could not be resolved on the basis of
response time data alone. In recent years, however, significant
efforts have investigated the neural underpinnings of the mental
rotation phenomena, arguing that the results of brain imaging studies
support the existence of visual imagery \cite{Kosslyn1988,
  Kosslyn:2009tj}.

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/stimuli_shapes.pdf}
    \vspace{0pt}
    \caption{Example stimuli}
    \label{fig:stimuli}
  \end{subfigure}
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/likelihood_function.pdf}
    \caption{Likelihood function}
    \label{fig:likelihood}
  \end{subfigure}
  \caption{\textbf{Rotated shapes and their similarity.}  \textbf{(a)}
    An example stimulus in which the shapes differ only by a
    rotation. All stimuli consist of three to six vertices centered
    around the origin, and edges which create a closed loop from the
    vertices. \textbf{(b)} The approximate likelihood (similarity) of
    $X_b$ given $X_R$, where $X_R$ is a rotation of $X_a$ by the angle
    $R$. The true angle of rotation between $X_a$ and $X_b$ is at
    $\frac{2\pi}{3}$, corresponding to the global maximum in the
    likelihood function.}
  \label{fig:shapes}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/li_regression.pdf}
    \caption{\Naive{} model}
    \label{fig:li}
  \end{subfigure}
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/vm_regression.pdf}
    \caption{Parametric model}
    \label{fig:vm}
  \end{subfigure}
  \caption{\textbf{Simple methods of estimating $S$.} In both cases,
    hill-climbing search is used until a maxima is found (in this
    case, at approximately $\frac{\pi}{6}$). The sampled points
    $\mathbf{R}$ (red circles) are then used to estimate $S$ (black
    lines are the true $S$, red lines are the estimate). \textbf{(a)}
    Linear interpolation. The overall estimate of $Z$ here will be too
    low, but certain angles will have a disproportionate contribution
    (e.g. between $\frac{\pi}{4}$ and $\frac{3\pi}{4}$. \textbf{(b)}
    Best fit of scaled Von Mises PDF parameters. As in (a), $Z$ will
    be underestimated. However, the fit around the maximum near
    $\frac{\pi}{6}$ is much more accurate.}
  \label{fig:simple-models}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/bq_regression.pdf}
  \caption{\textbf{Nonparametric model.} Each panel shows one step of
    the Bayesian Quadrature regression. Upper left: the original
    Gaussian Process (GP) regression for $S$. Lower left: the GP
    regression for $\log(S+1)$. Lower right: the GP regression for
    $\Delta_c=\mu_{\log S} - \log \mu_S$. Upper right: the adjusted
    regression for $S$, where the mean is equal to
    $\mu_S(1+\mu_{\Delta_c})$. The model uses this final estimate to
    compute $Z$ and will continue rotating until the variance of $Z$
    is low enough that a hypothesis may be accepted. This method
    allows the model to avoid local maxima such as the one near
    $\frac{\pi}{6}$, which causes trouble for the \naive{} and
    parametric models in Figure \ref{fig:simple-models}.}
  \label{fig:bq}
\end{figure}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/model_rotations.pdf}
  \caption{\textbf{Model rotations.} Top: each subplot shows the
    correspondence between the true angle of rotation ($R$) for
    ``same'' pairs and the amount of rotation performed by the
    model. Bottom: each subplot shows the models' mean rotations over
    stimuli pairs that were judged correctly. Black dots correspond to
    ``same'' pairs, and blue lines correspond to ``different'' pairs
    (for which the mean rotation is constant across true rotations, as
    the true rotation is undefined). Error bars/shaded regions
    indicate one standard deviation, and the dotted lines indicate the
    least-squares fit to the ``same'' pairs.}
  \label{fig:rotations}
\end{figure}

@@

The \naive{} model's response error rate was $\NaiveME{}$, which is
better than chance (equivalent to guessing randomly,
i.e. $\ME{}=0.5$). Closer inspection reveals that much of this comes
from ``different'' pairs ($\NaiveMEdiff{}$), where the asymmetric
linear interpolation may give an overestimate of $Z$ (e.g. Figure
\ref{fig:li}).  This intuition is confirmed by the model's accuracy in
estimating $Z$, which was $\NaiveMSE{}$. As shown in the left panel of
Figure \ref{fig:accuracy}, it overestimated $Z$ for nearly all stimuli
pairs.

@@

The parametric model's error in determining whether the objects were
identical was $\VMME{}$, making more errors on ``same'' pairs
($\VMMEsame{}$) than ``different'' pairs ($\VMMEdiff{}$). Rather than
overestimating as with the \naive{} model, however, the parametric
model largely underestimated $Z$. This is unsurprising: because the
Von Mises distribution only has a single peak, it necessarily
underestimates $Z$ for multimodal similarity functions. Overall, the
parametric model had an error rate of $\VMMSE{}$ in estimating $Z$,
which was significantly more accurate than that of the \naive{} model.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{../figures/Z_accuracy.pdf}
  \caption{\textbf{Accuracy in estimating $Z$.} Each subplot shows the
    true (``gold standard'') value of $Z$ vs. the value estimated by
    the model. Black dotted lines indicate a perfect 1:1
    correspondence between the true and estimated values. The \naive{}
    model (left panel) tends to overestimate. The parametric model
    (center panel) is very accurate for about half the ``same'' pairs,
    and severely underestimates the rest. The nonparametric model
    (right panel) maintains a decent correspondence with the true
    $Z$.}
  \label{fig:accuracy}
\end{figure}

@@

The nonparametric model was much more accurate in choosing the correct
hypothesis than the other two models ($\BQME{}$). We note that this is
very close to the 3.2\% error rate reported by \cite{Shepard1971},
though further experimentation is necessary to determine whether the
types of errors people make align with those of the model.

@@

% two questions: which way to rotate, when to stop, giving formal answer
% to these questions which produces answers consistent with human
% behavior. instance of showing people are appropriately using
% computationl resources
